---
title: "DP_MB_N_I_P_phyloseq"
author: "Apsara"
date: "2024-10-07"
output: html_document
---

This code has been modified by IP (10 Oct 2024) to include filtering, pruning, and rarefying of the phyloseq object prior to saving it as DP_phyloseq_object_final.RData

It also produces a DP_tax_matrix.RData file containing taxonomic information in a more readable format.

The resulting files are called in the DP_LogRatio_Work.Rmd code.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(phyloseq)
library(tidyr)
library(readr)
```

##Load the files and make phyloseq object

```{r phyloseq-object}
metafp <-"sample_manifest_updated.txt"
meta <- read_delim(metafp, delim="\t")

meta <- separate(meta, samplename, into = c("experiment", "mouse", "day", "group"), sep = "_")
meta<- subset(meta, day=="D0" & (group=="N"|group=="C"|group=="I"| group=="M"))

otufp <- "feature-table.txt"
otu <- read_delim(file= otufp, delim="\t", skip=1)

taxfp <- "taxonomy.tsv"
tax <- read_delim(taxfp, delim="\t")

phylotreefp <- "tree.nwk"
phylotree <-read_tree(phylotreefp)

otu_mat <- as.matrix(otu[,-1])
rownames(otu_mat) <-otu$'#OTU ID'
OTU <- otu_table(otu_mat, taxa_are_rows=TRUE)

samp_df <-as.data.frame(meta[,-1])
rownames(samp_df)<-meta$`#SampleID`
SAMP <-sample_data(samp_df)

tax_mat <- tax %>% select(-Confidence)%>% 
  separate(col=Taxon, sep="; ", into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")) %>%
  as.matrix()
tax_mat <-tax_mat[,-1]
rownames(tax_mat)<-tax$`Feature ID`
TAX <-tax_table(tax_mat)

object <-phyloseq(OTU, SAMP,TAX,phylotree) 
object_filt <- subset_taxa(object, Kingdom == "d__Bacteria" & Order!="o__Chloroplast" & Family !="f__Mitochondria")
object_filt_samples <- prune_samples(sample_sums(object_filt)>100, object_filt)
phyloseq_object_final <- rarefy_even_depth(object_filt_samples, rngseed = 1, sample.size =50000)

sample_data(object)
save(phyloseq_object_final,file="DP_phyloseq_object_final.RData")
```

### Create a taxonomy matrix
```{r tax-matrix}
## Now make tax_matrix into a data frame with OTU ID information
rownames(tax_mat) <- tax$`Feature ID`
tax_matrix <- tax_mat %>% as.data.frame()
tax_matrix$Genus <- gsub("g__", "", tax_matrix$Genus)
tax_matrix$Family <- gsub("f__", "", tax_matrix$Family)
tax_matrix$'#OTU ID' <- tax$`Feature ID`
colnames(tax_matrix)[colnames(tax_matrix) == 'Kingdom'] <- 'Domain'

save(tax_matrix,file="DP_tax_matrix.RData")
```