---
title: "LogRatio work"
author: "IP"
date: "2024-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
load("phyloseq_object_final.RData")
load("tax_matrix.RData") 

#Fix weird kingdom issue in taxmatrix

library(phyloseq) 
library(dplyr)
library(metagMisc)
library(ggplot2)
library(tidyr)
library(ComplexHeatmap)
library(microViz)
library(textshape)
library(circlize)
library(ggpubr)
```

### Overview
The goal here is to calculate the log-ratio change of certain taxa. The proposed strategy is as follows:

1. Find the family for which relative abundance remains ~constant throughout treatments

    - Lactobacillus, or bacteroides
    
2. Glom by that taxonomic level to get the absolute counts of that group (x)

3. For each group at that level, divide absolute counts by this value (abs/x) = ratio

4. Take log of this, and proceed to graphs

5. Use Aldex2 for statistical comparison

## Part 0 - convert phyloseq object to dataframes and calculate relative abundance
```{r setup-dataframes}
#Function to calculate relative abundance
calc_relative_abunance <- function(data) {
  data <- data %>%
    group_by(Sample) %>%
    mutate(Relative_Abundance = Abundance / sum(Abundance)) %>%
    ungroup()
  return(data)
}
#Columns to remove
unneeded <- c("forward.absolute.filepath","reverse.absolute.filepath","...8","...9")

#Convert phyloseq object to a dataframe for easier dplyr manipulation.
#Calculate relative abundance using our function.
phyloseq_df <- phyloseq_object_final %>% 
  psmelt() %>%
  select(-unneeded) %>%
  calc_relative_abunance()

#Do the same at family and genus level
phyloseq_fam_df <- phyloseq_object_final %>% 
  tax_glom(taxrank="Family") %>%
  psmelt() %>%
  mutate(Family = gsub("^f__", "", Family)) %>%
  select(-unneeded) %>%
  calc_relative_abunance()

phyloseq_genus_df <- phyloseq_object_final %>% 
  tax_glom(taxrank="Genus") %>%
  psmelt() %>%
  mutate(Genus = gsub("^g__", "", Genus)) %>%
  select(-unneeded) %>%
  calc_relative_abunance()
```

## Part 1 - identify stable families

```{r find-stable-families}
#To identify "stable" groups as a reference point (low variance between treatments and days).
abundance_threshold = 0.01 #threshold for minimum abundance values 
variance_threshold = 0.1 #threshold for maximum variance in abundance across samples

stable_fams <- phyloseq_fam_df %>%
    group_by(Family) %>%
    filter(all(Relative_Abundance > abundance_threshold)) %>%
    summarize(Variance = var(Relative_Abundance),
              Average_RA = mean(Relative_Abundance), 
              .groups = 'drop') %>%
    filter(Variance < variance_threshold) %>%
    select(Family,Average_RA,Variance)

stable_genera <- phyloseq_genus_df %>%
    group_by(Genus) %>%
    filter(all(Relative_Abundance > abundance_threshold)) %>%
    summarize(Variance = var(Relative_Abundance),
              Average_RA = mean(Relative_Abundance), 
              .groups = 'drop') %>%
    filter(Variance < variance_threshold) %>%
    select(Genus,Average_RA,Variance)

stable_fams
stable_genera
```


#### Cheeky visualisation
```{r visualise-stable-families}
#Define the families of interest, for now leaving their names ugly
families_oi <- stable_fams$Family
genera_oi <- stable_genera$Genus
stat_comparisons <- list(c("Mb_only","MB_Clostridia"),
                         c("Mb_only","Clostridia_only"),
                         c("Clostridia_only","MB_Clostridia"))
#Graph relative abundance between treatment groups.
ggplot(data = (phyloseq_fam_df %>% filter(Family %in% families_oi)),
        aes(x=Treatment, y=Relative_Abundance)) +
  geom_boxplot() +
  labs(title="Stable Families") + 
  facet_wrap(~Family) 

#Graph relative abundance between treatment groups.
ggplot(data = (phyloseq_genus_df %>% filter(Genus %in% genera_oi)),
        aes(x=Treatment, y=Relative_Abundance)) +
  labs(title="Stable Genera") +
  geom_boxplot()+
  stat_compare_means(comparisons = stat_comparisons, 
                     label = "p.signif", 
                     method = "t.test") +
  facet_wrap(~Genus)
```

## Part 2/3/4 Generate log-ratio values using a reference family/genus.
```{r logRatio-calculations}
# Define our reference family and genus.
ref_fam <- "Bacteroidaceae"
ref_gen <- "Bacteroides"
colours <- c("#FAA613","#F44708","#A10702")

#Mutate dataframes to calculate the abundance ratio and logratio, within samples.
#For 0 values, replace with LOI (half the lowest detected abundance in that sample)
phyloseq_fam_df <- phyloseq_fam_df %>% 
  filter(day==21) %>% #filter to just endpoint
  group_by(Sample) %>%
  mutate(min_abundance = min(Abundance[Abundance > 0], na.rm = TRUE),
         # Abundance = ifelse(Abundance == 0, min_abundance / 2, Abundance),
         Ratio = Abundance/Abundance[Family==ref_fam],
         LogRatio = log(Ratio)) %>%
  filter(Ratio!=0)

phyloseq_genus_df <- phyloseq_genus_df %>% 
  filter(day==21) %>% #filter to just endpoint
  group_by(Sample) %>%
  mutate(min_abundance = min(Abundance[Abundance > 0], na.rm = TRUE),
         # Abundance = ifelse(Abundance == 0, min_abundance / 2, Abundance),
         Ratio = Abundance/Abundance[Genus==ref_gen],
         LogRatio = log(Ratio)) %>%
  filter(Ratio!=0)

#Create filtered dataframes: contain only taxa which at least 2 treatments have measured abundance
phyloseq_genus_df_filtered <- phyloseq_genus_df %>% 
  ungroup() %>%  
  group_by(Genus) %>%
  filter(n_distinct(Treatment) == 3) %>%  
  ungroup()
phyloseq_fam_df_filtered <- phyloseq_fam_df %>% 
  ungroup() %>%  
  group_by(Family) %>%
  filter(n_distinct(Treatment) ==3 ) %>%  
  ungroup()

x <- ggplot(phyloseq_fam_df_filtered, aes(x=Treatment, y=LogRatio, fill=Treatment)) +
    geom_boxplot() + facet_wrap(~Family)  + 
    labs(title="Families") +
    scale_fill_manual(values=colours) +
    stat_compare_means(comparisons = stat_comparisons, vjust=0.1,
                     label = "p.signif", 
                     method = "t.test", 
                     bracket.size = 0.3,
                     y.position=1) 
x

x2 <- ggplot(phyloseq_genus_df_filtered, aes(x=Treatment, y=LogRatio, fill=Treatment)) +
    geom_boxplot() + facet_wrap(~Genus)+ 
    labs(title="Genera") +
    scale_fill_manual(values=colours) +
    stat_compare_means(comparisons = stat_comparisons, vjust=0.1,
                     label = "p.signif", 
                     method = "t.test", 
                     bracket.size = 0.3,
                     y.position=1) 
x2
```
One thing to note: Muribaculaceae doesn't show up because it is not detected in any of the Clostridia_only samples (which makes sense!) and it is thus filtered out for these boxplots.

### Repeating but as a for loop
Saves each individual boxplot as a pdf in the logratio_plots directory. Takes a second to run though.
```{r}
all_genus <- unique(phyloseq_genus_df$Genus)
all_fams <- unique(phyloseq_fam_df$Family)
colours <- c("#FAA613","#F44708","#A10702")

#Plot all genus graphs
for(g in all_genus){
  filtered <- phyloseq_genus_df %>% filter(Genus==g)
  p <- filtered %>% ggplot(aes(x = Treatment, y = LogRatio, fill = Treatment)) +
    geom_boxplot() +
    labs(title = paste("Boxplot for Genus:", g)) +
    scale_fill_manual(values=colours) +
    theme_minimal()
  plotpath <- file.path("logratio_plots/genus", paste0(g, ".pdf"))
  # ggsave(plotpath, plot = p, width = 8, height = 6)
}
#Plot all family graphs
for(f in all_fams){
  filtered <- phyloseq_fam_df %>% filter(Family==f)
  p <- filtered %>% ggplot(aes(x = Treatment, y = LogRatio, fill = Treatment)) +
    geom_boxplot() +
    labs(title = paste("Boxplot for Family:", f)) +
    scale_fill_manual(values=colours) +
    theme_minimal()
  plotpath <- file.path("logratio_plots/family", paste0(f, ".pdf"))
  # ggsave(plotpath, plot = p, width = 8, height = 6)
}

```

### Generation of simple heatmap
```{r}
# Create the heatmap
phyloseq_genus_df <- phyloseq_genus_df %>%
  filter(Family!="f__Chloroplast") %>%
  filter(Genus!="uncultured")
p <- ggplot(phyloseq_genus_df, 
            aes(x = Sample, y = Genus, fill = LogRatio)) +
  geom_tile() +
  labs(title = "Mean Log Ratio of Genus by Treatment",
       x = "Genus",
       y = "Treatment") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Generation of complex heatmap
Currently re-starting with the data to see if we can wrangle it as needed.
```{r}
load("phyloseq_object_final.RData")
ref_fam <- "Lactobacillaceae"
ref_gen <- "Bacteroides"

phyloseq_fam_df <- phyloseq_object_final %>% 
    tax_glom(taxrank="Family") %>%
    psmelt() %>%
    filter(day==21) %>%
    mutate(Family = gsub("^f__", "", Family)) %>%
    filter(Family!="uncultured") %>%
    calc_relative_abunance() %>%
  group_by(Family) %>%
  filter(!(Abundance==0)) %>% ungroup() %>%
   group_by(Sample) %>%
    mutate(min_abundance = min(Abundance[Abundance > 0], na.rm = TRUE),
           Abundance = ifelse(Abundance == 0, 0.5, Abundance),
           Ratio = Abundance/Abundance[Family==ref_fam],
           LogRatio = log(Ratio))  %>%
  ungroup() %>%  
  group_by(Family) %>%
  filter(n_distinct(Treatment) >1 ) %>%  
  ungroup()

for_heatmap <- phyloseq_fam_df %>%
  select(Sample, OTU, Family, LogRatio) %>%
  spread(key=Sample, value=LogRatio, fill=-10)  %>%
  column_to_rownames("Family") %>%
  select(-OTU) %>%
  as.matrix()

cols <- colorRamp2(c(1.0, 0.5, 0, -1, -2, -3, -4,-8,-10), 
                   c("#00441b", "#006d2c", "#238b45", "#41ae76", "#66c2a4", 
                     "#99d8c9", "#ccece6", "#e5f5f9", "#f7fcfd"))

#Design annotation to add the three groups as colorblocks
treatment_annotation <- HeatmapAnnotation(  
  df = data.frame(Treatment = factor(rep(1:3, each = 4), levels = 1:3)),
  annotation_legend_param = list(
    title = "Treatment",
    at = 1:3,
    labels = c("MB", "MB_Clostridia", "Clostridia")),
  col = list(Treatment = c("1" = "#FAA613", "2" = "#F44708", "3" = "#A10702")),
  annotation_height = unit(c(1), "cm")  )

h <- Heatmap(for_heatmap,
        name="LogRatio \nAbundance",
        row_names_side="left",
        column_names_side="top",
        show_row_names = TRUE,
        row_names_gp = gpar(fontsize = 8),
        show_column_names = FALSE,
        cluster_rows = TRUE,
        cluster_columns = FALSE,
        col=cols,
        column_title=paste("Normalised against",ref_fam),
        top_annotation = treatment_annotation)
h
# #Save the graph
# png("test.png",width=20,height=20,units="cm",res=1200)
# draw(h)
# dev.off()

# #Add this code to h to split the columns by treatment
        # column_split = factor(c("MB","MB","MB","MB",
        #                  "MB+Clos","MB+Clos","MB+Clos","MB+Clos",
        #                  "Clos","Clos","Clos","Clos"),
        #                  levels=c("MB","MB+Clos","Clos")),

```

```{r random trash}
# log_ratio_df <- phyloseq_fam_df %>%
#     select(Sample, Family, LogRatio) %>%
#     distinct()  # Ensure unique combinations
# 
# sample_data_df <- log_ratio_df %>%
#     pivot_wider(names_from = Family, values_from = LogRatio, values_fill = NA)
# 
# # Merge LogRatios into the phyloseq object
# # phyloseq_for_heatmap <- merge_phyloseq(
# #     tax_glom(phyloseq_object_final,taxrank="Family"),
# #     sample_data(sample_data_df)
# # )
# 
# phyloseq_for_heatmap <- phyloseq_object_final %>% tax_fix(unknowns = c("f__uncultured")) %>%
#   tax_transform(trans = "clr", rank = "Family")
# 
# 
# 
# taxa <- sample(tax_top(phyloseq_for_heatmap, n = 40), size = 20)
# 
# comp_heatmap(phyloseq_for_heatmap,
#              taxa=taxa)
```