---
title: "16S-processing"
author: "IP"
date: "2024-07-10"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(phyloseq)
library(ape)
library(tidyverse)
library(picante)
library(ggplot2)
library(ggpubr)
library(vegan)
library(DESeq2)
library(microViz)
```

## Make phyloseq object
```{r,echo=FALSE}
## import metadata and compress as needed
metafp <- "sample_manifest.txt"
meta <- read_delim(metafp, delim="\t") 
SAMP <- as.data.frame(meta[,-1])
rownames(SAMP) <- meta$`#SampleID`
SAMP <- SAMP %>% sample_data()

## import OTUs and make matrix
otufp <- "feature-table.txt"
otu <- read_delim(file = otufp, delim="\t", skip=1)
otu_mat <- as.matrix(otu[,-1]) 
rownames(otu_mat) <- otu$`#OTU ID`
OTU <- otu_table(otu_mat, taxa_are_rows = TRUE) 
class(OTU) 

## import taxa and make matrix
taxfp <- "2a_feature-table-with-taxonomy.tsv"
tax <- read_delim(taxfp, delim="\t", skip=1) 
tax_matrix <- tax %>% select(taxonomy) %>%
  separate(col=taxonomy, sep="; ",
	into=c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")) %>%  
  as.matrix()
rownames(tax_matrix) <- tax$`#OTU ID`
TAX <- tax_matrix %>% tax_table()

## Now make tax_matrix into a data frame with OTU ID information
tax_matrix <- tax_matrix %>% as.data.frame()
tax_matrix$Genus <- gsub("g__", "", tax_matrix$Genus)
tax_matrix$Family <- gsub("f__", "", tax_matrix$Family)
tax_matrix$'#OTU ID' <- tax$`#OTU ID`

## Create phylogenetic tree
phylotreefp <- "tree.nwk"
phylotree <- read.tree(phylotreefp)

## Create phyloseq object and filter out non-bacterial sequences
object <- phyloseq(OTU,SAMP,TAX,phylotree)
```

## Filter data (rarefy, etc.) and save phyloseq object.
```{r}
object_filt <- subset_taxa(object, Domain == "d__Bacteria" & Class!="c__Chloroplast" & Family !="f__Mitochondria")
object_filt_samples <- prune_samples(sample_sums(object_filt)>100, object_filt)

## Rarefy
# rarecurve(t(as.data.frame(otu_table(object_filt_samples))), cex=0.1)
phyloseq_object_final <- rarefy_even_depth(object_filt_samples, rngseed = 1, sample.size =50000)
phyloseq_object_final <- subset_samples(phyloseq_object_final, !sample_names(phyloseq_object_final) %in% c("MBClos-MBClos-inoc", "MBClos-Clos-inoc"))

## Save
save(object_filt_samples,file="object_filt_samples.RData")
save(phyloseq_object_final,file="phyloseq_object_final.RData")

```

## Perform alpha diversity analyses
```{r}
load("phyloseq_object_final.RData")

## Calculate Alpha diversity
alpha_div_data <- estimate_richness(phyloseq_object_final, measures = c("Observed", "Chao1", "Shannon")) %>% as.data.frame()
rownames(alpha_div_data) <- gsub("\\.", "-", rownames(alpha_div_data))
alpha_div_data$Sample <- rownames(alpha_div_data) 

## Factor metadata into our interests
meta_subset <- SAMP[,c("Treatment","day")] 
meta_subset$Sample <- rownames(SAMP)
meta_subset <- meta_subset[!is.na(meta_subset$Treatment),]
meta_subset$Treatment <- meta_subset$Treatment %>% factor( levels=c("Mb_only","MB_Clostridia","Clostridia_only")) 
meta_subset$day <- factor(meta_subset$day,levels=c("1","2","3","4","6","7","14","21"))

alpha_div_data <- alpha_div_data %>%
  left_join(meta_subset, by="Sample") %>% drop_na() 

alpha_div_data_long <- pivot_longer(alpha_div_data,
                                    cols = c(Observed, Chao1, Shannon),
                                    names_to = "Metric",
                                    values_to = "Value")

## Plot faceted by treatment and diversity index
ggplot(alpha_div_data_long, aes(x = day, y = Value, fill = Treatment)) +
  geom_boxplot(alpha = 0.3) +
  facet_grid(cols = vars(Treatment), rows = vars(Metric),scales = "free_y") +  # facet by both Treatment and Metric
  labs(x = "Day", y = "Diversity Metric") + 
  theme_linedraw()

## Plot just at day 21 with statistical comparisons (t-test)
alpha_div_data_long %>% filter(day==21) %>%
  ggplot(aes(x = Treatment, y = Value, fill = Treatment)) +
  geom_boxplot(alpha = 0.3) +
  scale_y_continuous(expand = expansion(mult = .2)) +
  facet_grid(rows = vars(Metric),scales = "free_y", switch="y") +  
  labs(x = "Treatment", y = "Diversity Metric") + 
  ggtitle("Alpha diversity after 21 days") +
  theme_linedraw() + 
  stat_compare_means(aes(group = Treatment), method = "t.test", 
                     comparisons = list(c("Mb_only", "MB_Clostridia"), c("Mb_only", 
                                    "Clostridia_only"), c("MB_Clostridia", "Clostridia_only")), 
                     label = "p.signif",step.increase = 0.2) 

alpha_div_data_long %>% filter(day==21) %>%
  ggplot(aes(x = Treatment, y = Value, fill = Treatment)) +
  geom_boxplot(alpha = 0.3) +
  scale_y_continuous(expand = expansion(mult = .2)) +
  facet_grid(rows = vars(Metric),scales = "free_y", switch="y") +  
  labs(x = "Treatment", y = "Diversity Metric") + 
  ggtitle("Alpha diversity after 21 days") +
  theme_linedraw() +
  geom_pwc(method="t_test", label="p.signif",hide.ns=TRUE)

```
## Perform beta diversity analyses
```{r}
# Goal: ß-diversity comparing between treatments at 1, 7, 14, and 21 days.
## Use phyloseq to calculate distance matrix
dm_weighted_unifrac <- UniFrac(phyloseq_object_final, weighted=TRUE)#weighted unifrac
dm_unweighted_unifrac <- UniFrac(phyloseq_object_final, weighted=FALSE)#unweighted unifrac
dm_bray <- vegdist(t(otu_table(phyloseq_object_final)), method="bray")#bray-curtis
dm_jaccard <- vegdist(t(otu_table(phyloseq_object_final)), method="jaccard")#jaccard

## For the timestamps given produce PCA plot coloured by Treatment
dayz <- c(1,7,14,21)
for (d in dayz){
  phyloseq_byday <- subset_samples(phyloseq_object_final,day==d)
  sample_data(phyloseq_byday)$Treatment <- factor(sample_data(phyloseq_byday)$Treatment, 
                                                     levels = c("Mb_only", "MB_Clostridia", "Clostridia_only"))
  ord.weighted_unifrac <- ordinate(phyloseq_byday, method="PCoA", distance="unifrac", weighted=TRUE)
  gg.weighted_unifrac <- plot_ordination(phyloseq_byday,
                                         ord.weighted_unifrac,color="Treatment") +
    labs(col="Treatment")+ stat_ellipse(type = "norm") +
    ggtitle(paste("Beta diversity",d,"days post treatment",sep=" ")) +
    theme_bw()
  print(gg.weighted_unifrac)
}
```
## #FUNCTION subset_my_phyloseq
###Takes in a phyloseq object and list of genera, returns an abundance table filtered.
```{r}
#Requires tax_matrix and otu_table in global environment
#Subsets by genus name
subset_my_phyloseq_genus <- function(physeq_object,genera){
  
  genus_names <- data.frame("Genus"=tax_matrix$Genus)
  rownames(genus_names) <- rownames(tax_matrix)
  genus_abundance <- tax_glom(physeq_object, taxrank = "Genus") %>%
    transform_sample_counts(function(x) x / sum(x)) 
  genus_abundance_table <- as.data.frame(otu_table(genus_abundance))
  rownames(genus_abundance_table) <- make.unique(genus_names[rownames(genus_abundance_table), 
                                                             "Genus"])
  ## Filter to species of interest
  filtered_abundance_table <- filter(genus_abundance_table, row.names(genus_abundance_table) %in%
                                       genera)
  ## Rework into usable data frame with metadata
  abundance_table_final <- filtered_abundance_table %>% t() %>% as.data.frame() %>%
    merge(meta_subset, by="row.names", all=FALSE)
  rownames(abundance_table_final) <- abundance_table_final[,1]
  abundance_table_final <- abundance_table_final[, -1] %>% 
    gather(Genus, Abundance, -Treatment, -day) %>%
    subset(Genus!="Sample")
  abundance_table_final$Abundance <- as.numeric(abundance_table_final$Abundance)
  return(abundance_table_final)
}

subset_my_phyloseq_fam <- function(physeq_object,families){
  fam_names <- data.frame("Family"=tax_matrix$Family)
  rownames(fam_names) <- rownames(tax_matrix)
  fam_abundance <- tax_glom(physeq_object, taxrank = "Family") %>%
    transform_sample_counts(function(x) x / sum(x)) 
  fam_abundance_table <- as.data.frame(otu_table(fam_abundance))
  rownames(fam_abundance_table) <- make.unique(fam_names[rownames(fam_abundance_table), 
                                                             "Family"])
  ## Filter to species of interest
  filtered_abundance_table <- filter(fam_abundance_table, row.names(fam_abundance_table) %in%
                                       families)
  ## Rework into usable data frame with metadata
  abundance_table_final <- filtered_abundance_table %>% t() %>% as.data.frame() %>%
    merge(meta_subset, by="row.names", all=FALSE)
  rownames(abundance_table_final) <- abundance_table_final[,1]
  abundance_table_final <- abundance_table_final[, -1] %>% 
    gather(Family, Abundance, -Treatment, -day) %>%
    subset(Family!="Sample")
  abundance_table_final$Abundance <- as.numeric(abundance_table_final$Abundance)
  return(abundance_table_final)
}
```
## Species-specific longitudinal analyses within treatments over time
## Genera lost in MB- mice
```{r}
# Goal: look at the following list of genera (lost in MB- mice) and count over time
## Genera producing butyrate, of interest
genera_of_interest <- c("Acetatifactor", "Anaerofustis", "Butyricicoccus", "Clostridia_UCG-014", "Lachnospiraceae_NK4B4_group", "Family_XIII_AD3011_group", "DNF00809", "Clostridia_vadinBB60_group", "Marvinbryantia", "Roseburia")

genera_of_interest <- TAX[TAX[, "Class"] == "c__Clostridia",]
genera_of_interest <- as.list(unique(genera_of_interest[, "Genus"]))
genera_of_interest <- gsub("g__", "", genera_of_interest) %>% as.list()

fam_of_interest <- TAX[TAX[, "Class"] == "c__Clostridia",]
fam_of_interest <- as.list(unique(fam_of_interest[, "Family"]))
fam_of_interest <- gsub("f__", "", fam_of_interest) %>% as.list()

## Make big plot with all days
# g <- ggplot(abundance_table_final, aes(x = factor(day), y = Abundance, fill = Treatment)) +
#   geom_boxplot() +
#   facet_grid(rows=vars(Genus),cols=vars(Treatment), scales = "free_y", switch="y") +
#   labs(x = "Day", y = "Relative abundance", title = "Relative Abundance of Genera by Treatment") +
#   theme_linedraw()

##PLOT 2 - Relative abundance of genera of interest (butyrates)
days=c(1,7,14,21)
for (d in days){
  temptitle <- paste("Relative abundance of genera on day",d)
  tempsave <- paste("plots/plot2/clostrid_abundance_day",d,".png",sep="")
  print(tempsave)
  g <- subset_my_phyloseq_genus(phyloseq_object_final,genera_of_interest) %>% filter(day==d) %>%
  ggplot(aes(x = Treatment, y = Abundance, fill = Treatment)) +
  geom_boxplot() +
    facet_wrap(vars(Genus), scales = "free_y") +
    labs(x = "", y = "Relative abundance", title = temptitle) +
    theme_linedraw() + 
    scale_y_continuous(expand = expansion(mult = .3)) +
    geom_pwc(method="t_test", label="p.signif",hide.ns=TRUE)+
    theme(axis.text.x = element_blank(),
        legend.position = "bottom") +
    expand_limits(y = 0)
  ggsave(filename = tempsave,width=12,height=7)
}

##PLOT3 - Repeated but for family level
days=c(1,7,14,21)
for (d in days){
  temptitle <- paste("Relative abundance of families on day",d)
  tempsave <- paste("plots/plot3/clostrid_fam_abundance_day",d,".png",sep="")
  print(tempsave)
  g <- subset_my_phyloseq_fam(phyloseq_object_final,fam_of_interest) %>% filter(day==d) %>%
  ggplot(aes(x = Treatment, y = Abundance, fill = Treatment)) +
  geom_boxplot() +
    facet_wrap(vars(Family), scales = "free_y") +
    labs(x = "", y = "Relative abundance", title = temptitle) +
    theme_linedraw() + 
    scale_y_continuous(expand = expansion(mult = .3)) +
    geom_pwc(method="t_test", label="p.signif",hide.ns=TRUE)+
    theme(axis.text.x = element_blank(),
        legend.position = "bottom") +
    expand_limits(y = 0)
  ggsave(filename = tempsave,width=12,height=7)
}
```

## Species-specific longitudinal analyses within treatments over time
## Look specifically at Lachnospiraceae family (PLOT1)
```{r}
# Goal: look at the following list of genera (genera in lachnospiraceae) and count over time

## Subset the phyloseq object to just include these genera
genera_of_interest <- TAX[TAX[, "Family"] == "f__Lachnospiraceae",]
genera_of_interest <- as.list(unique(genera_of_interest[, "Genus"]))
genera_of_interest <- gsub("g__", "", genera_of_interest) %>% as.list()

## Make big plot
g <- ggplot(subset_my_phyloseq(phyloseq_object_final,genera_of_interest), aes(x = factor(day), y = Abundance, fill = Treatment)) +
  geom_boxplot() +
  facet_grid(rows=vars(Genus),cols=vars(Treatment), scales = "free_y", switch="y") +
  labs(x = "Day", y = "Relative abundance", title = "Relative Abundance of Lachnospiraceae Genera by Treatment") +
  theme_linedraw()
g

## PLOT1: facet by day and genera on x.
days=c(1,7,14)
for (d in days){
  temptitle <- paste("Relative abundance of Lachnospiraceae on day",d)
  tempsave <- paste("plots/plot1/lachnospiraceae_abundance_day",d,".png",sep="")
  print(tempsave)
  g <- abundance_table_final%>% filter(day==d) %>%
  ggplot(aes(x = Treatment, y = Abundance, fill = Treatment)) +
  geom_boxplot() +
    facet_wrap(vars(Genus), scales = "free_y") +
    labs(x = "", y = "Relative abundance", title = temptitle) +
    theme_linedraw() + 
    scale_y_continuous(expand = expansion(mult = .3)) +
    geom_pwc(method="t_test", label="p.signif",hide.ns=TRUE)+
    theme(axis.text.x = element_blank(),
        legend.position = "bottom") +
    expand_limits(y = 0)
  print(g)
  ggsave(filename = tempsave,width=12,height=7)
}
```


#NOTES & TRASH
```{r}   
# ## Make DESEq object and add taxa information
# deseq_object <- phyloseq_to_deseq2(phyloseq_object_final, ~Treatment) %>% DESeq()
# deseq_results <- deseq_object %>% results(tidy=TRUE)
# colnames(deseq_results)[1] <- "#OTU ID"
# deseq_results_taxa <- inner_join(tax_matrix,deseq_results, by = "#OTU ID" )
# 
# ## Filter DESEq for significance p<0.01 and large changes.
# deseq_results_sig <-  deseq_results_taxa %>%
#   filter( padj<0.01 & abs(log2FoldChange)>2)
# deseq_results_sig <- deseq_results_sig[order(deseq_results_sig$log2FoldChange),]

#Make the plot
# ggplot(data = deseq_results_sig, aes(y = reorder(Genus, -(as.numeric(log2FoldChange))), x = log2FoldChange, fill = pvalue))+
# #   geom_col()
# 
# ## Count hits
# upregulated_count <- sum(deseq_results_sig$log2FoldChange > 0)
# downregulated_count <- sum(deseq_results_sig$log2FoldChange < 0)
# cat("Number of upregulated ASVs:", upregulated_count, "\n")
# cat("Number of downregulated ASVs:", downregulated_count, "\n")

//////
  
  # 
# ## Calculate all relative abundances
# genus_abundance <- tax_glom(phyloseq_object_final, taxrank = "Genus") %>%
#   transform_sample_counts(function(x) x / sum(x)) 
# genus_abundance_table <- as.data.frame(otu_table(genus_abundance))
# rownames(genus_abundance_table) <- make.unique(genus_names[rownames(genus_abundance_table), "Genus"])
# 
# ## Filter to species of interest
# filtered_abundance_table <- filter(genus_abundance_table, row.names(genus_abundance_table) %in% genera_of_interest)
# 
# ## Rework into usable data frame with metadata
# abundance_table_final <- filtered_abundance_table %>% t() %>% as.data.frame() %>%
#   merge(meta_subset, by="row.names", all=FALSE)
# rownames(abundance_table_final) <- abundance_table_final[,1]
# abundance_table_final <- abundance_table_final[, -1] %>% 
#   gather(Genus, Abundance, -Treatment, -day) %>%
#   subset(Genus!="Sample")
# abundance_table_final$Abundance <- as.numeric(abundance_table_final$Abundance)

subset_my_phyloseq_taxa <- function(physeq_object, taxa_list, level) {
  # Extract the taxonomy table from the phyloseq object
  tax_matrix <- tax_table(physeq_object)
  
  # Create a data frame of the taxonomic level of interest
  level_names <- data.frame("Taxa" = as.character(tax_matrix[, level]))
  rownames(level_names) <- rownames(tax_matrix)
  
  # Aggregate the OTU table at the specified taxonomic level
  taxa_abundance <- tax_glom(physeq_object, taxrank = level) %>%
    transform_sample_counts(function(x) x / sum(x))
  
  # Convert the aggregated OTU table to a data frame
  taxa_abundance_table <- as.data.frame(otu_table(taxa_abundance))
  rownames(taxa_abundance_table) <- make.unique(as.character(level_names[rownames(taxa_abundance_table), "Taxa"]))
  
  # Filter the abundance table to include only the taxa of interest
  filtered_abundance_table <- taxa_abundance_table %>% filter(row.names(.) %in% taxa_list)
  
  # Rework the filtered abundance table into a usable data frame with metadata
  abundance_table_final <- filtered_abundance_table %>% 
    t() %>% 
    as.data.frame() %>%
    merge(meta_subset, by = "row.names", all = FALSE)
  
  # Clean up the final data frame
  rownames(abundance_table_final) <- abundance_table_final[, 1]
  abundance_table_final <- abundance_table_final[, -1] %>%
    gather(Taxa, Abundance, -Treatment, -day) %>%
    subset(Taxa != "Sample")
  
  abundance_table_final$Abundance <- as.numeric(abundance_table_final$Abundance)
  return(abundance_table_final)
}
```   






